Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-30 analysis/paper/paper.Rmd
added SG gastric lavage edits

We examined the stomach contents of `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow from Deadwater Slough using gastric lavage methods. Of those, we found contents of some form in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) of those. Table \@ref(tab:gastric-tab) summarizes gastric lavage results for each sampling effort. For those individuals with stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r prettyNum(min(stmch_df$stomach_contents_weight), digits = 3)`-`r prettyNum(max(stmch_df$stomach_contents_weight), digits = 3)` g). Overall, fish or fish remnants consisted of `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall of all stomach contents examined.

@@

Across both seasons and all diet scenarios, we estimated the total number of fish consumed to be between `r prettyNum(round(min(total_impacts$smolt_N), 0), big.mark = ",")` and `r prettyNum(round(max(total_impacts$smolt_N), 0), big.mark = ",")` (Table \@ref(tab:adult-imp)). The estimated number of fish consumed by a Northern Pikeminnow begins to taper off between the 30-90% diet scenarios. So although we are uncertain of the diet proportions consisting of fish versus invertebrates, the impact of that uncertainty is somewhat constrained.

<!--
## Juvenile Chinook Salmon Abundance
For context, throughout the migration periods of DSR and NRR life histories it is estimated that 113,976 individuals are produced and therefore available within the upper Salmon River basin (Table x).

@@

The majority of fish caught during the study received gastric lavage. To evaluate the efficacy of the method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. In each case, no additional stomach contents were identified, suggesting that gastric lavage was effective in removing stomach contents, although our sample size was low. However, in total, we only identified fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`%) of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow examined using gastric lavage. Aside from juvenile Chinook Salmon, identified species included Redshine Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; however, decomposition from digestion rendered most stomach contents unidentifiable.

Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading...

<!--
Concerns with gastric lavage. We only observed juvenile Chinook salmon (and/or other fishes) in a very small number of stomach contents. However, we don't necessarily believe that to mean that pikeminnow in Deadwater Slough rarely consume juvenile Chinook salmon. Are there other cases where gastric lavage failed to document predation? How soon after consumption do you need to take a sample? So instead, we made some assumptions about the proportion of a pikeminnows diet consisting of Chinook salmon.
-->

>>>>

We examined the stomach contents of Northern Pikeminnow from Deadwater Slough using gastric lavage. To validate the efficacy of the method, we euthanized 75 Northern Pikeminnow after gastric lavage was completed and removed the remaining food items via dissection. We found that gastric lavage successfully removed food items in all dissected individual, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in some form in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) of those. Northern Pikeminnow captured during the spring 2021 effort had a higher proportion of diets with fish contents (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Spring_2021"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Spring_2021"], 1)`%) than fish captured in the fall 2019 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2019"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2019"], 1)`%) and fall 2020 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2020"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2020"], 1)`%) (Table \@ref(tab:gastric-tab)). For those individuals with stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r round(min(stmch_df$stomach_contents_weight), 2)`-`r round(max(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants were `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents examined.

@@

Across both seasons and all diet scenarios, we estimated the total number of fish consumed to be between `r prettyNum(round(min(total_impacts$smolt_N), 0), big.mark = ",")` and `r prettyNum(round(max(total_impacts$smolt_N), 0), big.mark = ",")` (Table \@ref(tab:adult-imp)). The estimated number of fish consumed by a Northern Pikeminnow begins to taper off between the 30-90% diet scenarios. So although we are uncertain of the diet proportions consisting of fish versus invertebrates, the impact of that uncertainty is somewhat constrained.
<!--
## Juvenile Chinook Salmon Abundance
For context, throughout the migration periods of DSR and NRR life histories it is estimated that 113,976 individuals are produced and therefore available within the upper Salmon River basin (Table x).

@@

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. Aside from juvenile Chinook Salmon, identified species included Redside Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; although decomposition from digestion rendered many fish contents unidentifiable. While this approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, previous studies have indicated that diets of angled fish differ from those collected by other methods (e.g., electrofishing, [@Hodgson1998]), and further, angled fish are much more likely to have empty stomachs [@Jurajda2016]. Second, unlike other predators such as Smallmouth Bass and Walleye that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Previous research indicates that performing gastric lavage on cyprinid species can be ineffective at quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Our intent, rather, was to simply document predation on juvenile Chinook Salmon in Deadwater Slough, and given the potential ineffectiveness of gastric lavage using angling and for Northern Pikeminnow, we consider our diet composition estimates to be conservative estimates i.e., we believe that a larger proportion of the Northern Pikeminnow diet is fish and juvenile Chinook Salmon than our results suggest. 
<!--
Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading...
-->
